Associations between chlorophyll *a* and various microcystin-LR health advisory

### 2016-01-15 manuscript/manuscript.tex
Wayne and Tim comments/edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Original default rstudio/pandoc latex file
%% upated by @jhollist 09/15/2014
%% inspired by @cboetting https://github.com/cboettig/template and
%% @rmflight blog posts:
%% http://rmflight.github.io/posts/2014/07/analyses_as_packages.html 
%% http://rmflight.github.io/posts/2014/07/vignetteAnalysis.html).  
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[11pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Associations between Chlorophyll a and various Microcystin-LR Health Advisory Concentrations},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Changes borrowed from @cboettig, added by @jhollist 
% A modified page layout 
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\usepackage{lineno} % add 
  \linenumbers % turns line numbering on 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%Packages and layout changes by @jhollist 09/15/2014
\usepackage{ragged2e}
\usepackage[font=normalsize]{caption}
  \usepackage[doublespacing]{setspace}
\usepackage{parskip}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\rfoot{\today}
\lfoot{\thepage}
%%Changed default abstract width and added lines
\renewenvironment{abstract}{
  \hfill\begin{minipage}{1\textwidth}
  \rule{\textwidth}{1pt}\vspace{5pt}
  \normalsize
  \begin{justify}
  \bfseries\abstractname\vspace{5pt}
  \end{justify}}
  {\par\noindent\rule{\textwidth}{1pt}\end{minipage}
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Associations between Chlorophyll \emph{a} and various Microcystin-LR
Health Advisory Concentrations}
\author{
Jeffrey W. Hollister
Betty J. Kreakie
}
\date{}

\begin{document}
%%Edited by @jhollist 09/15/2014
%%Adds title from YAML
\begin{singlespace}
\begin{center}
\huge Associations between Chlorophyll \emph{a} and various Microcystin-LR
Health Advisory Concentrations
\end{center}
%%Adds Author, correspond email asterisk, and affilnum from YAML
\begin{center}
\large
Jeffrey W. Hollister \textsuperscript{*} \textsuperscript{1} 
Betty J. Kreakie \textsuperscript{1} 
\end{center}
%%Adds affiliations from YAML
\begin{justify}
\footnotesize \emph{ 
\\*
\textsuperscript{1}US Environmental Protection Agency, Office of Research and Development,
National Health and Environmental Effects Research Laboratory, Atlantic
Ecology Division, 27 Tarzwell Drive Narragansett, RI, 02882, USA\\*
}
%%Adds corresponding author email(s) from YAML
\newcounter{num}
\setcounter{num}{1}
\\[0.1cm]
\footnotesize \emph{ 
\ifnum\value{num}=1%
\textsuperscript{*} corresponding author:
\fi
\href{mailto:hollister.jeff@epa.gov}{\nolinkurl{hollister.jeff@epa.gov}}
\stepcounter{num}
}
\end{justify}
%%Adds date from YAML
\normalsize

\end{singlespace}

@@

\vspace{2mm}

\hrule

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. In particular,
chlorophyll \emph{a} is very commonly measured, can be estimated
\emph{in situ}, and has been shown to be positively associated with
microcystin. In this paper, we use this association to provide estimates
of chlorophyll \emph{a} concentrations that if exceeded would be
indicative of a higher probability of exceeding select health advisory
concentrations for microcystin-LR. Using the 2007 National Lakes
Assessment and a conditional probability approach, we identify
chlorophyll \emph{a} concentrations that are more likely than not to be
associated with an exceedance of a microcystin health advisory level. We
look at the recent US EPA health advisories for drinking water as well
as the World Health Organization levels for drinking water and
recreational use. For the specific advisory microcystin concentrations
of 0.3, 1, 1.6, and 2, we find chlorophyll \emph{a} concentrations of
0.07, 0.07, 2.79, and 11.36, respectively. When managing for these
various microcystin levels, exceeding these reported chlorophyll
\emph{a} concentrations should be a trigger for further testing and
possible management action.

\vspace{3mm}

\hrule

\doublespace

\section{Introduction}\label{introduction}

Over the last decade or so numerous events and legislative activities
have raised the public awareness of harmful algal blooms (Rinta-Kanto et
al. 2009, HABHRCA 2014, Jetoo et al. 2015), and in response the US
Environmental Protection Agency (USEPA) has recently released suggested
microcystin-LR (one of the more common toxins) concentrations that would
trigger health advisories (McElhiney and Lawton 2005, Zurawell et al.
2005, USEPA 2015). Additionally, the World Health Organization (WHO) has
had proposed advisory levels for drinking water and a range of
recreational risk levels (Chorus and Bartram 1999, Organization 2003).
While these levels and associated advisories are likely to help mitigate
the impacts from harmful algal blooms, they are not without
complications.

One of these complications is that they rely on available measurements
of microcystin-LR. While laboratory testing remains the gold standard
for quantifying microcystin-LR concentrations in water samples, several
field test kits have been developed. Even though field tests provide a
much needed means for rapid assessment, they are not yet widely used and
are moderately expensive (approximately \$150-\$200 depending on
specific kit) with a limited shelf life (typically one year) (James et
al. 2011, Aranda-Rodriguez et al. 2015). Additionally, each technique
requires nuanced understanding of the detection method (e.g., limit of
detection, specific microcystin variants being measured, and sampling
protocol).

Fortunately, microcystin-LR has been shown to be associated with several
other, more commonly measured and well understood components of water
quality that are readily assessed in the field. For instance, there are
small or hand held fluorometers that measure chlorohpyll \emph{a}.
Additionally, chlorophyll \emph{a} is a very commonly measured component
of water quality that is also known to be positively associated with
microsystin-LR concentrations (Pip and Bowman 2014, Yuan et al. 2014).
Yuan et. al (2014) explore these associations in detail and control for
other related variables. In their analysis they find that total nitrogen
and chlorophyll \emph{a} show the strongest association with
microcystin. Furthermore, they identify chlorophyll\emph{a} and total
nitrogen concentrations that are associated with exceeding 1 \(\mu\)g/L
of microcystin. Given these facts, it should be possible to identify
chlorophyll \emph{a} concentrations that would be associated with the
new USEPA Microcystin-LR health advisory levels for drinking water.
Identifying these associations would provide another tool for water
resource managers to help manage the threat to public health posed by
cHABs and would be especially useful in the absence of microcystin-LR
concentrations.

In this paper we build on past efforts and utilize the National Lakes
Assessment (NLA) data and identify chlorophyll \emph{a} concentrations
that are associated with higher probabilities of exceeding several
microcystin-LR health advisory concentrations (Chorus and Bartram 1999,
USEPA 2009, 2015). We add to past studies by exploring associations with
newly announced advisory levels and by also applying a different method,
conditional probability analysis. Utilizing different methods
strengthens the evidence for suggested chlorophyll \emph{a} levels that
are associated with increased risk of exceeding the health advisory
levels as those levels are not predicated on a single analytical method.
So that others may repeat or adjust this analysis, the data, code, and
this manuscript are freely available via
\href{https://github.com/USAPE/microcystinchla}{\url{https://github.com/USAPE/microcystinchla}}.

\section{Methods}\label{methods}

\subsection{Data}\label{data}

We used the 2007 NLA water quality and microcystin-LR concentration data
(USEPA 2009). These data represent a snapshot of water quality from the
summer of 2007 for the conterminous United States. Data on chlorophyll
\emph{a} and microcystin-LR concentrations are available for lakes.

\subsection{Conditional Probability
Analysis}\label{conditional-probability-analysis}

We used a conditional probability analysis (CPA) approach to explore
associations between chlorophyll \emph{a} concentrations and World
Health Organization (WHO) and USEPA microcystin-LR health advisory
levels (Paul and Munns 2011). Many levels have been suggested (Table
\ref{tab:microcystin_levels}), but lakes with higher microcystin-LR
concentrations in the NLA were rare. Only 1.16 \% of lakes sampled had a
concentration greater than 10. Thus, for this analysis we focus on the
microcystin concentrations that are better represented in the NLA data.
These were the USEPA children's drinking water advisory level of 0.3
\(\mu\)g/L (USEPA Child), the WHO drinking water advisory level of 1
\(\mu\)g/L (WHO Drinking), the USEPA adult drinking water advisory level
of 1.6 \(\mu\)g/L (USEPA Adult), and the WHO recreational, low
probability of effect advisory level of 2 \(\mu\)g/L (WHO Recreational).

Conditional probability analysis provides information about the
probability of observing one event given another event has also occured.
For this analysis, we used CPA to examine how the conditional
probability of exceeding one of the health advisories changes as
chlorophyll \emph{a} increases in a lake. We expect to find higher
chlorohpyll \emph{a} concentrations to be associated with higher
probabilities of exceeding the microcystin-LR health advisory levels. We
also caclulated bootstrapped 95\% confidence intervals (CI) using 1000
bootstrapped samples. Thus, to identify chlorophyll \emph{a}
concentrations of concern we identfiy the value of the upper 95\% CI
across a range of conditional probabilities of exceeding each health
advisory level. As both microcystin-LR and chlorophyll \emph{a} values
were highely skewed right, a log base 10 transformation was used.
Additional details of the specific implementation are available at
\url{https://github.com/USEPA/microcystinchla}. A more detailed
discussion of CPA is beyond the scope of this paper, but see Paul et al.
(2005) and Hollister et al. (2008) for greater detail.

\section{Results}\label{results}

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225
\(\mu\)g/L. Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent
the detection limits. Any value greater than that indicates the presence
of microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51 and the mean was 3.17. Of all lakes sampled, 21\%
of lakes exceeded the USEPA Child level, 8.8\% of lakes exceeded the
USEPA Adult level,11.7\% of lakes exceeded the WHO Drinking level,and
7.3\% of lakes exceeded the WHO Recreational level. For chlorophyll
\emph{a}, the range was 0.07 to 936 \(\mu\)g/L. All lakes had reported
chlorophyll \emph{a} concentrations that exceeded detection limits. The
median concentration was 7.79 \(\mu\)g/L and the mean was 29.63
\(\mu\)g/L. The associations between chlorophyll \emph{a} and the upper
confidence interval across a range of conditional probability values is
shown in Table \ref{tab:mc_chla_table}. Specific chlorophyll \emph{a}
that are associated with greater than even odds of exceeding the
advisory levels were 0.07, 0.07, 2.79, and 11.36 for 0.3, 1.0, 1.6 and
2.0 \(\mu\)g/L advisory levels, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

\section{Discussion}\label{discussion}

The association between Log10 microcystin-LR and Log10 chlorophyll
\emph{a} shows a wedge pattern (Figure \ref{fig:chla_micro_scatter}).
This indicates that higher concentrations of microcystin-LR almost
always co-occur with higher concentrations of chlorophyll \emph{a} yet
the inverse is not true. Higher chlorophyll \emph{a} is not necessarily
predictive of higher microcystin-LR concentrations; however, chlorophyll
\emph{a} may be predictive of the probability of exceeding a certain
concentration.

This is the case as the probability of exceeding each of the four tested
health advisory levels increases as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We use this association
to identify chlorophyll \emph{a} concentrations that are associated with
greater than even odds of exceeding a given health advisory level (Table
\ref{tab:mc_chla_table}). These represent 99.9\%, 99.9\%, 77\%, and
43.8\% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA
Adult, and the WHO recreational levels, respecitvely.

Furthermore, the chlorophyll \emph{a} cutoffs may be used to predict
whether or not a lake exceeds the microcystin-LR health advisories.
Doing so allows us to compare the accuracy of the prediction as well as
evaluate false negatives. Total accuracy of the four cutoffs predicting
microcystin-LR exceedances were 21\% for the USEPA children's advisory,
12\% for the WHO drinking water advisory, 31\% for the USEPA adult
advisory and, 61\% for the WHO regreational advisory (Tables
\ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}).
However, total accuracy is only one part of the prediction performace
with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin-LR exceedances, the error that should be avoided is
predicting that no exceedance has occurred when in fact it has. In other
words, we would like to avoid Type II errors and minimize the proportion
of false negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 0\%, 0\%, 0\% and , 1\% for the U.S EPA
childrens drinking water, the WHO drinking water, the USEPA adult
drinking water, and the WHO recreational advisories, respectively. In
each case we miss less than 10\% of the lakes that are in fact exceeding
the microcystin-LR advisory.

There are numerous possible uses for the chlorophyll \emph{a} and
microcystin-LR advisory cut-off values. First, in the absence of
microcystin-LR measurements, exceedence of the chlorophyll \emph{a}
concentrations could be a trigger for further actions. Given that there
is uncertainity around these chlorophyll \emph{a} cutoffs the best case
scenario would be to monitor for chlorophyll \emph{a} and in the event
of exceeding a target concentration take water samples and have those
samples tested in a lab for microcystin-LR.

A second potential use is to identify possible bloom events from
historical data. As harmful algal blooms are made up of many species and
have various mechanisms responsible for adverse impacts (e.g.~toxins,
hypoxia, odors), there is no single definition of a bloom. For cHABs one
approach has been to identify an increase over a baseline concentration
of phycocyanin (Miller et al. 2013). This is a useful approach for
targeted studies, but phycocyanin is also not always available and
measures the predominance of cyanobacterial pigments and not toxins.
Using our chlorophyll \emph{a} cutoffs provides a value that is more
directly associated with microcystin-LR and can be used to classify
lakes, from past surveys, as having bloomed.

Lastly, using chlorophyll \emph{a} is not meant as a replacement for
testing of microcystin-LR. It should be used when other, direct
measurements of cyanotoxins are not available. In those cases, which are
likely to be common at least in the near future, using a more ubiquitous
measurement, such as chlorophyll \emph{a} will provide a reasonable
proxy for the probability of exceeding a microcystin-LR health advisory
level and provide better protection against adverse effects in both
drinking and recreational use cases.

\section{Acknowledgements}\label{acknowledgements}

@@

\section{Figures}\label{figures}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/chla_micro_scatter-1.pdf}
\caption{Scatterplot showing association betweeen chlorophyll \textit{a}
and microcystin-LR. \label{fig:chla_micro_scatter}}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin-LR (MLR) health advisory
Levels. A.) Plot for USEPA Child (0.3 \(\mu\)g/L). B.) Plot for WHO
Drinking (1 \(\mu\)g/L). C.) Plot for USEPA Adult (1.6 \(\mu\)g/L). D.)
Plot for WHO Recreational (2 \(\mu\)g/L). \label{fig:multi_cp_plot}}
\end{figure}

@@

\section{Tables}\label{tables}

>>>>

---
title: Associations between chlorophyll *a* and various microcystin-LR health advisory
  Concentrations
author:
- affilnum: 1
  email: hollister.jeff@epa.gov
  name: Jeffrey W. Hollister
- affilnum: 1
  name: Betty J. Kreakie
affiliation:
- affil: US Environmental Protection Agency, Office of Research and Development, National
    Health and Environmental Effects Research Laboratory, Atlantic Ecology Division,
    27 Tarzwell Drive  Narragansett, RI, 02882, USA
  affilnum: 1
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    template: components/f1000research.latex
  html_document: null
capsize: normalsize
csl: components/ecology.csl
documentclass: article
fontsize: 11pt
linenumbers: yes
bibliography: components/manuscript.bib
spacing: doublespacing
abstract: no
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Microsystin Chlorophyll Manusript}
-->

@@

\vspace{2mm}\hrule

<!-- Abstract is being wrapped in latex here so that all analysis can be run in the chunk above and the results reproducibly referenced in the abstract. -->
Cyanobacteria harmful algal blooms (cHABs) are associated with a wide range of adverse health effects that stem mostly from the presence of cyanotoxins.  To help protect against these impacts, several health advisory levels have been set for some toxins.  In particular, one of the more common toxins, microcystin, has several advisory levels set for drinking water and recreational use.  However, compared to other water quality measures, field measurements of microcystin are not commonly available due to cost and advanced understanding required to interpret results.  Addressing these issues will take time and resources.  Thus, there is utility in finding indicators of microcystin that are already widely available, can be estimated quickly and *in situ*, and used as a first defense against high levels of microcystin.  Chlorophyll *a* is commonly measured, can be estimated *in situ*, and has been shown to be positively associated with microcystin.  In this paper, we use this association to provide estimates of chlorophyll *a* concentrations that are indicative of a higher probability of exceeding select health advisory concentrations for microcystin-LR.  Using the 2007 National Lakes Assessment and a conditional probability approach, we identify chlorophyll *a* concentrations that are more likely than not to be associated with an exceedance of a microcystin health advisory level.  We look at the recent US EPA health advisories for drinking water as well as the World Health Organization levels for drinking water and recreational use and identify a range of chlorophyll *a* thresholds.  A 50% chance of exceeding one of the specific advisory microcystin concentrations of 0.3, 1, 1.6, and 2 $\mu$g/L is associatied with chlorophyll *a* concentration thresholds of 22.72, 77.2, 84.96, and 89.71, respectively.  When managing for these various microcystin levels, exceeding these reported chlorophyll *a* concentrations should be a trigger for further testing and possible management action.

\vspace{3mm}\hrule
\doublespace

#Introduction

Over the last decade, numerous events and legislative activities have raised the public awareness of harmful algal blooms [@jetoo2015toledo; @rinta2009lake; @HABHRCA2014]. In response the US Environmental Protection Agency (USEPA) has recently released suggested microcystin-LR (one of the more common toxins) concentrations that would trigger health advisories [@mcelhiney2005detection; @zurawell2005hepatotoxic; @usepa2015drinking].  Additionally, the World Health Organization (WHO) has proposed microcystin advisory levels for drinking water and a range of recreational risk levels [@world2003cyanobacterial; @chorus1999toxic].   While these levels and associated advisories are likely to help mitigate the impacts from harmful algal blooms, they are not without complications.

One of these complications is that they rely on available measurements of microcystin-LR.  While laboratory testing (e.g., chromatography) remains the gold standard for quantifying microcystin-LR concentrations in water samples, several field test kits have been developed.  Even though field tests provide a much needed means for rapid assessment, they are not yet widely used and are moderately expensive (approximately $150-$200 depending on specific kit) with a limited shelf life (typically one year) [@james2011environmental; @aranda2015evaluation].  Additionally, each technique requires nuanced understanding of the detection method (e.g., limit of detection, specific microcystin variants being measured, and sampling protocol).

Fortunately, microcystin-LR has been shown to be associated with several other, more commonly measured and well understood components of water quality that are readily assessed in the field.  For instance, there are small or hand held fluorometers that measure chlorohpyll *a*.  Additionally, chlorophyll *a* is a very commonly measured component of water quality that is also known to be positively associated with microsystin-LR concentrations [@pip2014microcystin; @yuan2014managing].  Yuan et. al [-@yuan2014managing] explore these associations in detail and control for other related variables.  In their analysis they find that total nitrogen and chlorophyll *a* show the strongest association with microcystin. Furthermore, they identify chlorophyll*a* and total nitrogen concentrations that are associated with exceeding 1 $\mu$g/L of microcystin.   Given these facts, it should be possible to identify chlorophyll *a* concentrations that would be associated with the new USEPA microcystin-LR health advisory levels for drinking water.  Identifying these associations would provide another tool for water resource managers to help manage the threat to public health posed by cHABs and would be especially useful in the absence of measured microcystin-LR concentrations.

In this paper we build on past efforts and utilize the National Lakes Assessment (NLA) data and identify chlorophyll *a* concentrations that are associated with higher probabilities of exceeding several microcystin-LR health advisory concentrations [@usepa2009national; @usepa2015drinking; @chorus1999toxic].  We add to past studies by exploring associations with newly announced advisory levels and by also applying a different method, conditional probability analysis.  Utilizing different methods strengthens the evidence for suggested chlorophyll *a* levels that are associated with increased risk of exceeding the health advisory levels as those levels are not predicated on a single analytical method.  So that others may repeat or adjust this analysis, the data, code, and this manuscript are freely available via [https://github.com/USEPA/microcystinchla](https://github.com/USAPE/microcystinchla).

#Methods

##Data
We used the 2007 NLA chlorophyll *a* and microcystin-LR concentration data [@usepa2009national, [NLA](http://www2.epa.gov/national-aquatic-resource-surveys/national-lakes-assessment)]. These data represent a snapshot of water quality from the summer of 2007 for the conterminous United States and were collected as part of an ongoing probabilistic monitoring program [@usepa2009national]. Data on chlorophyll *a* and microcystin-LR concentrations are available for  lakes.

##Analytical Methods
We used a conditional probability analysis (CPA) approach to explore associations between chlorophyll *a* concentrations and World Health Organization (WHO) and USEPA microcystin-LR health advisory levels [@paul2011probability].  Many health advisory levels have been suggested (Table \ref{tab:microcystin_levels}), but lakes with higher microcystin-LR concentrations in the NLA were rare. Only 1.16% of lakes sampled had a concentration greater than 10 $\mu$g/L.  Thus, for this analysis we focused on the microcystin concentrations that are better represented in the NLA data. These were the USEPA children's drinking water advisory level of  0.3 $\mu$g/L (USEPA Child), the WHO drinking water advisory level of 1 $\mu$g/L (WHO Drinking), the USEPA adult drinking water advisory level of 1.6 $\mu$g/L (USEPA Adult), and the WHO recreational, low probability of effect advisory level of 2 $\mu$g/L (WHO Recreational).

Conditional probability analysis provides information about the probability of observing one event given another event has also occured.  For this analysis, we used CPA to examine how the conditional probability of exceeding one of the health advisories changes as chlorophyll *a* increases in a lake.  We expect to find higher chlorohpyll *a* concentrations to be associated with higher probabilities of exceeding the microcystin-LR health advisory levels.  We also calculated bootstrapped 95% confidence intervals (CI) using 1000 bootstrapped samples.  Thus, to identify chlorophyll *a* concentrations of concern we identfiy the value of the upper 95% CI across a range of conditional probabilities of exceeding each health advisory level.  Using the upper confidence limit to identify a threshold is justified as it ensures that a given threshold is unlikely to miss a microcystin exceedance.

As both microcystin-LR and chlorophyll *a* values were highely skewed right, a log base 10 transformation was used.  Additional details of the specific implementation are available at https://github.com/USEPA/microcystinchla.  A more detailed discussion of CPA is beyond the scope of this paper, but see Paul et al. [-@paul2005development] and Hollister et al. [-@hollister2008cprob] for greater detail.  Lastly, all analyses were conducted using R version 3.2.2 and code and data from this analysis are freely available as an R package at [https://github.com/USEPA/microcystinchla](https://github.com/USAPE/microcystinchla).

Lastly, we assess the ability of these chlorophyll *a* thresholds to predict microcystin exceedance.  We use error matrices and calculate total accuracy as well as the proportion of false negatives.  Total accuracy is the total number of correct predictions divided by total observations.  The proportion of false negatives is the total number of lakes that were predicted to not exceed the microcystin guidelines but actually did, divided by the total number of observations.

#Results
In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225 $\mu$g/L.  Microcystin-LR concentrations of 0.05 $\mu$g/L represent the detection limits.  Any value greater than that indicates the presence of microcystin-LR.  Of those lakes with microcystin, the median concentration was 0.51$\mu$g/L and the mean was 3.17$\mu$g/L.  Of all lakes sampled, 21% of lakes exceeded the USEPA Child level, 8.8% of lakes exceeded the USEPA Adult level, 11.7% of lakes exceeded the WHO Drinking level, and 7.3% of lakes exceeded the WHO Recreational level. For chlorophyll *a*, the range was 0.07 to 936 $\mu$g/L.  All lakes had reported chlorophyll *a* concentrations that exceeded detection limits.  The median concentration was 7.79 $\mu$g/L and the mean was 29.63 $\mu$g/L.  The associations between chlorophyll *a* and the upper confidence interval across a range of conditional probability values are shown in Table \ref{tab:mc_chla_table}.  Specific chlorophyll *a* that are associated with greater than even odds of exceeding the advisory levels were 0.07, 0.07, 2.91, and 12.53$\mu$g/L for 0.3, 1.0, 1.6, and 2.0 $\mu$g/L advisory levels, respectively (Table \ref{tab:mc_chla_table} & Figure \ref{fig:multi_cp_plot}).

The chlorophyll *a* cutoffs may be used to predict whether or not a lake exceeds the microcystin-LR health advisories.  Doing so allows us to compare the accuracy of the prediction as well as evaluate false negatives.  Total accuracy of the four cutoffs predicting microcystin-LR exceedances were 19% for the USEPA children's advisory, 22% for the WHO drinking water advisory,  22% for the USEPA adult advisory, and 23% for the WHO regreational advisory (Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table}, \ref{tab:adult_conmat_table}, & \ref{tab:who_rec_conmat_table}).  However, total accuracy is only one part of the prediction performace with which we are concerned.

When using the chlorophyll *a* cutoffs as an indicator of microcystin-LR exceedances, the error that should be avoided is predicting that no exceedance has occurred when in fact it has.  In other words, we would like to avoid Type II errors and minimize the proportion of false negatives.  For the four chlorophyll *a* cut-offs we had a proportion of false negatives of 2%, 2%, 1%, and 1% for the USEPA children's, the WHO drinking water, the USEPA adult, and the WHO recreational advisories, respectively.  In each case we missed less than 10% of the lakes that in fact exceeded the microcystin-LR advisory.  While this method performs well with regard to the false negative percentage, it is possible that is a relic of the NLA dataset and testing with additional data would allow us to confirm this result.

#Discussion
The association between Log10 microcystin-LR and Log10 chlorophyll *a* shows a wedge pattern (Figure \ref{fig:chla_micro_scatter}).  This indicates that, in general, higher concentrations of microcystin-LR almost always co-occur with higher concentrations of chlorophyll *a* yet the inverse is not true.  Higher chlorophyll *a* is not necessarily predictive of higher microcystin-LR concentrations; however, chlorophyll *a* may be predictive of the probability of exceeding a certain threshold.

This is the case as the probability of exceeding each of the four tested health advisory levels increases as a function of chlorophyll *a* concentration (Figure \ref{fig:multi_cp_plot}).  We used this association to identify chlorophyll *a* concentrations that are associated with a range of probabilities of exceeding a given health advisory level (Table \ref{tab:mc_chla_table}).  For the purposes of this discussion we focus on a conditional probability of 50% or greater (i.e., greater than even odds to exceed a health advisory level). The 50% conditional probability chlorophyll *a* thresholds represents 28.7%, 9.8%,
8.6%, and 8.3% of sample lakes for the USEPA Child, the WHO Drinking, the USEPA Adult, and the WHO recreational levels, respecitvely.

There are numerous possible uses for the chlorophyll *a* and microcystin-LR advisory cut-off values.  First, in the absence of microcystin-LR measurements, exceedence of the chlorophyll *a* concentrations could be a trigger for further actions.  Given that there is uncertainity around these chlorophyll *a* cutoffs the best case scenario would be to monitor for chlorophyll *a* and in the event of exceeding a target concentration take water samples and have those samples tested for microcystin-LR.

A second potential use is to identify past bloom events from historical data.  As harmful algal blooms are made up of many species and have various mechanisms responsible for adverse impacts (e.g., toxins, hypoxia, odors), there is no single definition of a bloom.  For cHABs, one approach has been to identify an increase over a baseline concentration of phycocyanin [@miller2013spatiotemporal].  This is a useful approach for targeted studies, but phycocyanin is also not always available and measures the predominance of cyanobacterial pigments and not toxins. Using our chlorophyll *a* cutoffs provides a value that is more directly associated with microcystin-LR and can be used to classify lakes, from past surveys, as having bloomed.

Lastly, using chlorophyll *a* is not meant as a replacement for testing of microcystin-LR or other toxins.  It should be used when other, direct measurements of cyanotoxins are not available.  In those cases, which are likely to be common at least in the near future, using a more ubiquitous measurement, such as chlorophyll *a* will provide a reasonable proxy for the probability of exceeding a microcystin-LR health advisory level and provide better protection against adverse effects in both drinking and recreational use cases.

#Acknowledgements
We would like to thank Anne Kuhn, Bryan Milstead, John Kiddon, Joe LiVolsi, Tim Gleason, and Wayne Munns for constructive reviews of this paper. This paper has not been subjected to Agency review. Therefore, it does not necessary reflect the views of the Agency. Mention of trade names or commercial products does not constitute endorsement or recommendation for use. This contribution is identified by the tracking number ORD-015143 of the Atlantic Ecology Division, Office of Research and Development, National Health and Environmental Effects Research Laboratory, US Environmental Protection Agency.

@@

#Figures

![Scatterplot showing association betweeen chlorophyll \textit{a} and microcystin-LR. \label{fig:chla_micro_scatter}](figure/chla_micro_scatter-1.pdf)

@@

<!--

```
## No id variables; using all as measure variables
```

![Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. \label{fig:chla_micro_cdf}](figure/chla_micro_cdf-1.pdf)

\newpage-->

```
## Loading required package: grid
```

![Conditional probability plots showing association between the probability of exceeding various microcystin-LR (MLR) health advisory Levels. A.) Plot for USEPA Child (0.3 $\mu$g/L). B.) Plot for WHO Drinking (1 $\mu$g/L). C.) Plot for USEPA Adult (1.6 $\mu$g/L). D.) Plot for WHO Recreational (2 $\mu$g/L). \label{fig:multi_cp_plot}](figure/epa_child_cp_plot-1.pdf)

@@

#Tables

|Source |Type                                    |Concentration    |
|:------|:---------------------------------------|:----------------|
|USEPA  |Adult Drinking Water Advisory           |1.6 $\mu$g/L     |
|USEPA  |Child Drinking Water Advisory           |0.3 $\mu$g/L     |
|WHO    |Drinking Water                          |1 $\mu$g/L       |
|WHO    |Recreational: High Prob. of Effect      |20-2000 $\mu$g/L |
|WHO    |Recreational: Low Prob. of Effect       |2-4 $\mu$g/L     |
|WHO    |Recreational: Moderate Prob. of Effect  |10-20 $\mu$g/L   |
|WHO    |Recreational: Very High Prob. of Effect |>2000 $\mu$g/L   |

:Various suggested microcystin-LR health advisory concentrations. \label{tab:microcystin_levels}
